---
title: "Calibration of Urban Planning and Transport Health Assessment Toolkit"
output:
   #word_document: default
   #github_document
   #pdf_document:
   #  number_sections: yes
   html_document:
        toc: true
        toc_float: true
        number_sections: true
        theme: flatly
bibliography: ../ATF.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
library(dplyr)
knitr::opts_chunk$set(
  echo = FALSE,
  out.width = "100%",
  collapse = TRUE,
  comment = "#>",
  fig.path="figures/"
)
if (!file.exists ("ATF.bib")) { # delete file to auto-update
    refs <- RefManageR::ReadZotero(group = "418217",
                    .params = list(collection = "TJEQMXJR", limit = 100))
    RefManageR::WriteBib(refs, "ATF.bib")
}
library (ggplot2)
theme_set (theme_minimal ())
```


# Introduction

Calibration of UPTHAT (the Urban Planning and Transport Health Assessment
Toolkit) relies primarily on new high performace software
[@padgham_dodgr:_2019] for analysing dynamic movement through urban areas.

This manuscript details procedures developed for calibrating UPTHAT estimates
of both pedestrian and cyclist densities along individual street segments
against empirical data from New York City, USA.

# Methods

## Data Sources

UPTHAT is based out of principle on entirely open data sets, primarily on data
from Open Street Map and "worldpop" (www.worldpop.org). The latter data enable
estimates of population flows to be made in absolute terms, given aggregate
data on proportions and typical frequencies of trips.

Additional data on pedestrian densities was obtained from New York City
Government [pedestrian count
data](https://www1.nyc.gov/html/dot/html/about/datafeeds.shtml#Pedestrians),
collected bi-annually at 114 stations throughout the city, and on cyclist
densities from data provided by the city's "citibike" public hire bicycle
system. Additional data on subway usage was obtained from the city's
[Metropolitan Transit
Authority (MTA)](http://web.mta.info/nyct/facts/ridership/ridership_sub_annual.htm),
with data on locations obtained through the city's [open data
portal](https://data.cityofnewyork.us/Transportation/Subway-Stations/arq3-7z49).
The MTA figures are annual "ridership totals" for each station, which are
counts of number of passengers entering the system at each of 424 stations (or,
in a few cases, "station complexes", at which a single entrance provides access
to multiple proximate stations and multiple lines). There are no data for
system exits, and so we assumed that exit numbers equalled entrance
numbers throughout. The station locations given by the City of New York are
single coordinates for each station (473 in total). The City also provides
coordinates of all subway *entrances* (1,926 in total), along with associated
subway lines to which those entrances lead, with entrances often providing
access to multiple subway lines. We allocated the ridership data to subway
entrances by associating each entrance for a given line with the station
coordinates for that line, and associating each subway entrance with the
nearest station for that line. Entrances were associated with individual
stations by repeating this procedure for all subway lines, resulting in many
entrances being associated with both different subway lines and different
stations, but nevertheless providing an accurate connection between subway
entrances, the stations to which they lead, and the MTA's ridership counts.
Counts for each station were then distributed equally among all associated
subway entrances, ultimately converting 424 counts into 1,926 locations of
associated pedestrian flows.



### The Street Network

Both the street network used to route pedestrian journeys, and the structure,
nature, and location of buildings serving as origins and/or destinations of
those journeys were obtained from Open Street Map, accessed via the R package,
`osmdata` [@padgham_osmdata_2017]. The network consisted of 813,402 street
segments, representing a total 30,226 km. These segments are simply those used
to represent Open Street Map, such that a single linear length of way may be
represented by several points, and several corresponding edges in our internal
representation. We then used the `dodgr` software [@padgham_dodgr:_2019] to
"contract" this network down to segments between junctions points only,
reducing it to 407,840 segments.

Each of these 407,840 segments had an associated length, along with further
detail as contained in the Open Street Map data, including the type of way
(with such values as "cycleway", "footway", "pedestrian", or "steps"). These
types were used to weight the distances to reflect typical pedestrian
preferences [@luxen_real-time_2011], to enable pedestrian routes to be
accurately calculated. These weighting schemes act to extend the effective
length of ways unsuitable for pedestrians, for example, by multiplying actual
lengths of major vehicular roads by some factor representative of the
associated pedestrian discomfort in comparison to quiet, pedestrian-only ways.
All weighting factors are greater than one, so that the most
pedestrian-friendly ways retain their original distances, while ways less
suitable for pedestrian travel have weighted lengths greater than their actual
lengths. Routing is then conducted using these weighted lengths, and so routes
reflect those that would be typically chosen by pedestrians. Rather than
distance-based routing, we used time-based routing, which operates on a similar
principle, but introduces several categories of additional time-based
penalties, notably for waiting at traffic lights, or for waiting to cross busy,
multi-lane roads (where such crossing is permitted). All routing was
implemented using the authors' own `dodgr` software [@padgham_dodgr:_2019].

### Trip categories

Open Street Map includes not only streets, but also buildings with optional yet
frequently extensive descriptions of categories, purpose, sizes, and other
properties. These building descriptions enable Open Street Map to be used to
identify likely locations of trip origins and destinations according to trip
purpose. In particular, we divided building categories into the five main
groups of transportation, sustenance, entertainment, education, and healthcare,
each by grouping the Open Street Map tags given in Appendix 1. Transportation
notably included all parking facilities, both for bicycles and automobiles.

For each of these categories, we estimated local intensities of the respective
activities by attributing all buildings to the nearest street network junction,
and aggregating numbers at each junction. These numbers provided estimates of
likely numbers of people originating from or coming to that point for the
respective activities, and are referred to throughout the present work as
"densities." For transportation, Open Street Map data commonly include
capacities of parking facilities, whether for bicycles or automobiles. These
capacities were taken as the densities of points which provided them, while all
other points were assigned the average capacity of all functionally similar
points (so, for example, bicycle parking facilities absent specified capacity
were assigned the average capacity of all bicycle parking facilities with
specified capacities).

Beyond these Open Street Map-based estimates, we also used Worldpop data to
ascribe aggregate residential densities to each street junction, and subway
counts to represent journeys to or from entrances and exits of the New York MTA
subway system. A final category was based on the betweenness centrality of the
entire street network. Centrality is known to be a strong determinant of travel
behaviour, with more central points, or street segments, correlated with higher
densities of movements towards or away from those points or segments. A final
categorical trip purpose considered in our analyses below is thus,
"centrality".

Extraction and processing of these data enabled analyses of the following eight
trip categories, serving as types of journey origins or destinations:

n | category
--- | ---
1 | subway
2 | centrality
3 | residential
4 | transportation
5 | sustenance
6 | entertainment
7 | education
8 | healthcare

Each category of trip represented one pairwise combination of these eight
categories, yielding 8&#215;7 = 56 distinct categories. An additional eight
categories were formed through modelling random dispersal away from spatial
locations identified for each category.

## Modelling Pedestrian Flows

We modelled pedestrian flows throughout the entirety of New York City, USA, in
order to calibrate our models against the empirical data from the 114
pedestrian count stations. Models were constructed using the
[`dodgr`](https://github.com/atfutures/dodgr) software package for large-scale
routing through street networks, by assembling an array of "flow layers"
representing pedestrian flows between the distinct categories of origin and
destination points described above.

Flow layers were generated in two primary ways: either by dispersing from
a given set of origin points according to specified densities, or through
routing flows between specific sets of origin and destination points, with
densities calculated by a doubly-constrained spatial interaction model, with
exponential form used throughout [@wilson_boltzmann_2008]. We refer to these
two methods of calculating flow layers as "flow dispersal" and "flow
aggregation", and illustrate both in the following sub-sections.



### Flow Layers

Our models comprised a variety of flow layers, each of which was generated by
applying the spatial interaction models described above between a specified set
of origin points, taken from the eight categories given above. Each flow layer
was obtained through calculating shortest paths between each origin point, $i$,
and destination point, $j$, in the network, and aggregating flows according to
exponential spatial interaction models [@wilson_boltzmann_2008] of the form,
\begin{equation}
    SI_{ij} (d_{ij}) = \frac{n_in_{j\ne i}}{\sum_{j\ne i} n_j exp (-d_{ij} / k)} exp (-d_{ij} / k),
\end{equation}
where $d_{ij}$ denotes the distance between the points $i$ and $j$ (and
self-flows $SI_{ii}$ were excluded). The denominator ensured that our models
were singly-constrained to unit sums for each origin, $i$, over densities at
destinations, $j$. The above form gives the expected flow, in direct units of
$n_i$, along the path between the points $i$ and $j$. Layers were also
calculated describing undirected dispersal throughout the entire network from
a set of origin points.

Our [`dodgr`](https://github.com/ATFutures/dodgr) software enables spatial
interaction models to be efficiently calculated for a range of exponential
decay coefficients, `k`, returning a matrix of flows, with one row for each
segment of the network, and one column for each exponential decay coefficient
entered. This enabled us to initially calculate all flow layers for defined
values of `k`, and to subsequently combine these pre-calculated layers in
a single model as described in the following section. Each layer was calculated
for 30 values of `k`, from 100 to 3000 metres in 100 metre increments.

Finally, the city-wide flows from these flow layers, amounting in each layer to
values along over 800,000 street segments, were mapped on to the locations of
the pedestrian count stations. Flow layers contained non-zero values for
the actual street segments on which count stations were located only where the
data describing the start and end points translated into at least one route
passing along that segment. This was not always the case, and so values for
each pedestrian counter were aggregated from a selected number of nearest
non-zero flow values. This number itself was varied between 1 and 20, and the
value chosen which gave the minimal error model following the procedures in the
subsequent section. This approach yielded flow values at pedestrian count
stations which were aggregated over potentially differing numbers of nearest
street segments for the different layers, but we argue that this procedure is
appropriate because actual placements of flow values for each layer depend on
the spatial location and variability of data as recorded in Open Street Map.


### Statitstical Models

The 64 pairwise combinations of origin and destination categories represents 64
potential independent variables in a statistical model. Our final model
incorporated only a small fraction of this total number, through applying
a step-wise variable addition procedure. The first layer was selected by
applying the same procedure to each layer of aggregated flows generated by each
of the 30 value of `k`, across numbers of nearest segments, `n`, from 1 to 20
as described above, to determine the values of `k` and `n` that minimised the
error of a standard linear regression against observed pedestrian counts. The
layer which yielded the overall minimal error was selected as the first model
layer.

The procedure was then repeated by applying an analogous procedure to all 63
remaining layers, and including the flow values from the previously-selected
layer in a multiple linear regression model. Having selected the second layer,
it was then also included along with each of the remaining 62 layers, in order
to determine the third layer yielding the model with the lowest overall error.
Flow layers were successively added to the model as long as their contribution
to the model was significant; that is, layer addition within the model was
terminated as soon as the next layer added made no significant contribution.

Where the addition of new layers rendered the contribution any any prior layers
no longer significant, those non-significant layers were removed from the model.
Finally, it is statistically possible that a flow layer is significantly
*negatively* correlated with observed counts. Such layers were not included
with our models, which were constructed exclusively from layers which were
positively correlated with observed counts.


## Calibration to Observed Values

The observed pedestrian counts are given as total numbers observed
within either two- or three-hour windows (twice on weekdays, between
7-9am and 4-7pm; and once on Saturdays, between 12-2pm). We used the
weekday counts to yield an overall aggregate estimate of numbers per
day. To scale the observed values, we compared them with values derived
from the United Kingdom’s National Travel Survey
([NTS](https://beta.ukdataservice.ac.uk/datacatalogue/series/series?id=2000037),
waves 2002-2017 representing 5.8m single stage journeys), from which we
obtained a nation-wide estimate of numbers of walking trips for each of the 24
hours of the day (see Figure 1).


```{r hourly-walking, fig.cap = "Figure 1: Hourly proportions of total daily walking trips as obtained from the UK's National Travel Survey, with shaded areas illustrating the hours of pedestrian counting start between 7-9am and 4-7pm"}
f <- "wh.csv"
if (!file.exists (f))
{
    u <- "https://github.com/ATFutures/ATF-ideas/releases/download/0.0.1/wh.csv"
    invisible (download.file (u = u, destfile = f))
}
x <- read.csv ("wh.csv")
x$hour <- as.integer (substring (x$hour, 1, 2))
ymax <- max (x$perc)
ggplot (x, aes (x = hour, y = perc)) +
    geom_rect (aes (xmin = 7, ymin = 0, xmax = 10, ymax = ymax),
               fill = "#66666605") +
    geom_rect (aes (xmin = 16, ymin = 0, xmax = 20, ymax = ymax),
               fill = "#66666605") +
    geom_line () +
    geom_point () +
    ylab ("hourly proportion of total daily walking trips")
```

```{r hourly-walking-stats, echo = FALSE}
index <- which (x$hour %in% c (7:9, 16:19))
walk_total <- signif (100 * sum (x$perc [index]), 3)
walk_prop <- signif (sum (x$perc [index]), 2)
walk_conversion <- signif (1 / sum (x$perc [index]), 3)
```


These data reflected a total proportion of trips between the hours of 7-9am and
4-7pm of `r walk_total`%. We assumed this figure to approximately reflect
global pedestrian behaviour, and so converted observed pedestrian counts to
expected equivalent daily totals for weekdays by multiplying by 1 / 
`r walk_prop` = `r walk_conversion`.




# Results

The flow layer which made the most significant initial contribution reflected
dispersal from subway stations. The model error along the 30 values of
exponential decay coefficients is illustrated in Figure 1, and was typical for
most layers, manifesting a clear and distinctive minimum.

```{r get-layer-data, echo = FALSE}
f <- file.path (here::here (), "calibration", "sub-cen-model.Rds")
dolayer <- !file.exists (f)
```


```{r layer-error-with-k-calc, eval = dolayer}
library (calibration)
data_dir <- "/data/data/moveability/nyc"
net <- readRDS (file.path (data_dir, "calibration", "net-sub-cen.Rds"))
p <- ped_osm_id (data_dir = data_dir, net = net, quiet = TRUE)
saveRDS (p, file = file.path (here::here (), "calibration", "p.Rds"))
dp <- dodgr::dodgr_dists (net, from = p$id)

# code from calibration::opt_fit_to_ped
fcols <- grep ("flow", names (net))
sorted_verts <- function (dmat)
{
    res <- array (NA, dim = dim (dmat))
    for (i in seq (nrow (dmat)))
    {
        res [i, ] <- colnames (dmat) [order (dmat [i, ])]
    }
    return (res)
}
dp_verts <- sorted_verts (dp)
# then convert that to equivalent indices into both .vx0 and .vx1 of net_f
index0 <- t (apply (dp_verts, 1, function (i) match (i, net$.vx0)))
index1 <- t (apply (dp_verts, 1, function (i) match (i, net$.vx1)))

flowmat <- as.matrix (net [, fcols])
n <- 5 # number of neighbours to aggregate
temp <- rcpp_match_flow_mats (flowmat, index0 - 1, index1 - 1, fcols - 1, n)
saveRDS (temp, file = f)
```
```{r layer-error-with-k}
p <- readRDS (file.path (here::here (), "calibration", "p.Rds"))
temp <- readRDS (f)
ss <- apply (temp, 2, function (j) {
                 mod <- summary (lm (p$week ~ j))
                 sum (mod$residuals ^ 2) / 1e6  })
dat <- data.frame (k = 1:30 * 100, ss = ss)
x0 <- dat$k [which.min (dat$ss)]
y0 <- min (dat$ss)
ggplot (dat, aes (x = k, y = ss)) +
    geom_line () +
    geom_point (aes (x = x0, y = y0), cex = 3) +
    xlab ("k (metres)") +
    ylab ("Model error (in observed daily counts)")
```




# References
