---
output: github_document
# output:
#   pdf_document:
#     number_sections: true
bibliography: ATF.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
remotes::install_github("robinlovelace/gitgantt")
library(dplyr)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
if (!file.exists ("ATF.bib"))
{
    refs = RefManageR::ReadZotero(group = "418217", .params = list(collection = "TJEQMXJR", limit = 100))
    RefManageR::WriteBib(refs, "ATF.bib")
}
```

# Introduction

This document sets out a plan for the third phase of work for the WHO's Urban Health Initiative, and the ongoing development of an "Active Transport Tool" [ATT; @lovelace_estimating_2018].
**The main output will be a prototype web tool -- the ATT -- to enable interactive viewing of current patterns of active transport and associated health impacts, as well as the effects of scenarios of change.**
The work will build on previous projects funded by the WHO (in grants 2017/773067 and 2017/773067-0).
The ATT can be used for many purposes related to the interface between transport and human health, including assessment of exposure to air pollution, changes in the distribution of physical activity, and the spatial distribution of cycling and pedestrian flow densities.
These flow densities --- quantified as relative densities of bi-directional movement along each segment of the urban street network --- are calculated for a number of "layers" reflecting distinct journey purposes and modes.

The driving aim of this project across all stages will be to create of a *flexible framework* for analyzing, visualizing and testing a range of scenarios based on different input datasets and parameters, and to ensure the framework is able to be updated and modified at any stage within or beyond the project timetable as desired.
This work will go beyond the Propensity to Cycle Tool (PCT) project [@lovelace_propensity_2017], which has become a part of legally binding legislation and is being used by dozens of transport planning organizations around the UK.
The longer term aim is to create a globally scalable tool for sustainable transport planning.

# Project stages

As in previous phases, this project will involve the development of reproducible code and the creation of open data, for future use.
The emphasis will shift from experimental software development and proof-of-concept production (delivered in phases 1 and 2) to the development of a stable prototype application.
As outlined in the next section, the 'stable application' will have a dedicated domain and be available for at least 2 months after completion of the project.
The main stages of the work will be:

<!-- 1. **Analytic Methods and Software** Phase II resulted in a prototype ATT representing patterns of active travel in Accra only.  -->
<!--    This third Phase will apply all previous analyses to Kathmandu, thereby consolidating code and ensuring transferability between locations. -->

1. **Health Impacts** This stage will involve converting metrics of mobility (number of movement along street segments for a range of journey purposes, modes of transport, and demographic factors) into health-economic measures, based on the Health Economic Assessment Tool [HEAT @kahlmeier_health_2014]  .
   The primary aim of this translation in to health impacts will be to enable comparison of scenarios under the fourth point below in health-economic terms, although we will also investigate the feasibility of using locally-provided data to calibrate a static HEAT model.
1. **Scenario development**  
1. **Prototype App** Concurrent with the preceding two phases, the prototype app will be set-up and served from a stable web location, and will be maintained for the duration of the project.
1. **Health Impacts of Scenarios** The output of the previous phase will be combined with those of the second phase to enable comparison of scenarios in terms of their impact on health-economic measures.
1. **User Manual** Final phases of this project stage will yield a user manual able to be taken to the study cities of Accra and Kathmandu, to enable local stakeholders to understand, utilise, and provide feedback on the tool.
   As stated above, the tool itself will be web-based, and this usual manual will presume as little computer expertise as possible, and should be intelligible to an entirely general audience.
1. **Adaptation Manual** Finally, we will also develop an "Adaptation Manual" serving the dual purpose of describing

   (i) How an ATT may be adapted and applied to other, additional locations; and in doing so,
   (ii) How the ATT as presented to each location may be adapted and modified following feedback from local stakeholders.
   
   This manual will detail in general, non-technical terms, the nature of inputs, analytic processes, and resultant outputs, to enable stakeholders to readily identify modifications that might ultimately provide outputs better suited to local needs, desires, and future visions.

# Deliverables

The primary deliverables extending from each of the above phases will be:

<!-- 1. **Analytic Methods and Software** This phase will not produce any concrete deliverables, but will provide for both cities the necessary input data to allow the development of individually adapted ATT. -->
1. **Health Impacts** This phase will enable ATT output to be expressed in health-economic terms.
   As the third phase will be conducted concurrently with these first two, the output of this phase will be directly viewable in the ATT.
1.  **Prototype App** The single deliverable of this phase will be the ATT, for which both cities will be able to be selected, and a variety of ATT outputs viewed. This will be hosted at a dedicated URL and be available for at least 2 months after the project is complete.
1.  **Scenarios** \<TODO: RL\> Plausible scenarios of change, including:
    - A global, citywide scenario of uptake of active transport, as a result of citywide policies to promote walking and cycling. This would use baseline levels of walking and cycling from the global south as a reference, and an uptake scenario such as Go Dutch for cycling.
    - A global, citywide scenario of multi-modal tranpsort change, showing reduced levels of driving following disincentives to own and use cars.
    - A global scenario of public transport uptake, linked to [SDG 11](https://sustainabledevelopment.un.org/sdg11).
    - Locally specific scenarios, such as creation of car-free zones or reductions in car parking spaces.
1. **Health Impacts of Scenarios** This phase will extend the App to include the ability to select and compare scenarios, in terms both of changes in mobility patterns as well as associated health and economic terms.
1. **User and Adaptation Manuals** The final two phases will deliver both the User and Adaptation manuals for the ATT.

# Resources and timeline


## Gantt chart

```{r gantt, echo=FALSE, message=FALSE}
issue_df = gitgantt::gg_issue_df(owner = "ATFutures", repo = "who3")
issue_df_min = issue_df %>% 
   filter(!grepl(pattern = "Analytic", x = title))
gitgantt::gg_timevis(issue_df_min)
```

# References
