---
title: "Health Impacts of Scenarios"
output:
   #word_document: default
   #github_document
   #pdf_document:
   #  number_sections: yes
    html_document:
        toc: true
        toc_float: true
        number_sections: true
        theme: flatly
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  out.width = "100%",
  collapse = TRUE,
  comment = "#>",
  fig.path="figures/"
)
```

# Adaptation of the WHO's Health and Economic Assessment Tool (HEAT) to spatially explicit analyses

We adapt HEAT methodology for scenario calibrations, using the general
(non-spatial) parameters described in the first of the following sections, with
spatially explicit adaptations of other parameters described in the subsequent
section. More specific details are given in the accompanying `"heat.Rmd"`
document.

## General parameters

We adopt the following general parameters directly from HEAT methodology. The
relative reduction in risk for walking is assumed to be 0.886, such that the effect
for walking a certain distance under a given scenario of
```}
0.114 * d-scenario / d-reference
```
where `d_reference` and `d_scenario` are respective distances walked under the
reference and scenario conditions. Risk reduction for walking is capped at 30%,
but does not approach that value in any of our models.

Each increment of exposure of +10 ug/m^3 (considered to refer to PM2.5
throughout, as in HEAT, with effects of other pollutants such as PM10 assumed
to be linearly related to effects of PM2.5) is assumed to increase all-cause
mortality risk by 1.07. Effects of increased exposure are presumed to cease above 
50ug/m^3. In contrast to HEAT, which assumes the increase in exposure for
walking to be a constant factor of 1.6, we use values derived from literature
specific to Accra, with spatial interpolation provided by our models as
described below.

In the absence of a specific general mortality rate for Accra, we simply
averaged values for all countries from the [WHO mortality
database](http://apps.who.int/healthinfo/statistics/mortality/whodpms). The
figures are included here, yielding a global average mortality rate as follows:
```{r who-mortality}
x <- read.csv ("./who-mortality.csv")
names (x) <- c ("country", "population", "deaths", "remove")
x$remove <- NULL
x <- x [!is.na (x$deaths), ]
mortality <- mean (x$deaths / x$population)
message ("global mortality = ", mortality)
```
The respective populations of Accra and Kathmandu are presumed to be is 2.27
and 1.74 million, as taken from from wikipedia.

## Accra-specific data 

### Air Pollution

An article on [air pollution in sub-Saharan African
cities](https://link.springer.com/article/10.1007%2Fs11869-013-0199-6) has an
appendix
[here](https://static-content.springer.com/esm/art%3A10.1007%2Fs11869-013-0199-6/MediaObjects/11869_2013_199_MOESM1_ESM.docx)
that includes figures for Accra from [this
paper](https://pubs.acs.org/doi/10.1021/es903276s). That estimates average
PM2.5 concentrations throughout the year *apart from* Dec-Jan, when dust blows
in from the Sahara, of 39-53ug/m^3 at roadside sites, and 30-70ug/m^3 at
residential sites. Across the whole year, average values at roadside sites were
80-108ug/m^3, and at residential sites 57-106ug/m^3.

The paper helpfully simply states that PM2.5 concentrations at roadside sites
were 8-14ug/m^3 higher at roadside than at residential sites. This suggests we
can assume a background value of 35ug/m^3, with road-borne emissions adding
another 14ug/m^3 at the highest vehicular intensities. The following analyses
assume these values, and allocate the additional 14ug/m^3 assumed to be caused
by vehicular pollution according to a standard model of automobile traffic
throughout the city. Vehicular emissions are presumed to disperse away from
direct linear (road-based) sources via a Gaussian dispersal model with an
assumed half-width of 200 metres. The following results are largely invariant
to precise choice of this dispersal width.

### Walking statistics

We derived estimates of average daily walking distance from  data previously
shared by WHO, as follows. Average distances walked to the nearest market (in
kilometres) were given as in the following table, which translates to an
average distance of 2.77 km:

<1      | 1.1-2   | 2.1-3 | 3.1-6   | 6.1-10
--- | --- | --- | --- | ---
27.3    | 21.2    | 6.1   | 42.4    | 3.0

Distances walked from home to the nearest bus (trotro) stop were similarly
tabulated as follows, translating to an average distance to trotro of 0.51 km:

0-0.5   | 0.6-1 | 1.1-2 | 2.1-5 | > 5
--- | --- | --- | --- | ---
83.2    | 11.9  | 2.3   | 0.5   | 2.2

We then estimated average distances walked to work from statistics for mode of
transport taken to work of 47.4% walking and 19% trotro. Walks to work were
assumed to reflect distances typical of the above walks to the nearest market,
giving a resultant mean distance to work of `d_work = (0.474 * d_work + 0.19
* d_trotro) / (0.474 + 0.19) = 2.12 km`

Numbers of trips per day was also tabulated as follows, translating to an
average of 11.2 walking trips per day.

0-10    | 11-20 | 21-30 | 31-40 | 41-60 | 61-100 | > 100
--- | --- | --- | --- | --- | --- | ---
64.0    | 20.4  | 6.2   | 2.9   | 3.4   | 0.7    | 0.0

This number of trips per day was converted to a weekly estimate of 78.8 trips.
Of these trips, an average of 2.5 per person per week were presumed to be trips
to work, and an equivalent number trips to the nearest markets (or other trips
over equivalent distances). The distances of one half of the remaining 73.8
trips were presumed to be typified by distances to trotro, and so have mean
values of 0.51 km, the remaining half to be double that distance. This yielded
an estimated average weekly distance walked of 66.3 km, or a daily value of
9.48 km. This figure may seem unusually high, especially compared with the
baseline value assumed in HEAT of 14.84 km *per week*, but it must also be
noted that 93% of all Accra residents walk to purchase food or other
necessities.

## Scenarios

The above figures can be translated into effective increase in distance walked
per week in response to a relative modal increases in walking. These increased
can then also be translated according to the standard HEAT values along with
the global average mortality rate, into an estimate of reduction in mortality
in Accra. The resultant figures are shown in the following graph.
```{r mode_shift, echo = FALSE}
x <- read.csv ("./who-mortality.csv")
names (x) <- c ("country", "population", "deaths", "remove")
x$remove <- NULL
x <- x [!is.na (x$deaths), ]
mortality <- mean (x$deaths / x$population)
mode_shift <- function (mode_incr = 0.01, accra_pop, mortality)
{
    d_market <- c (0.5, 1.5, 2.5, 4.5, 8.5)
    p_market <- c (0.273, 0.212, 0.061, 0.424, 0.03)
    d_market <- sum (d_market * p_market)

    d_tro <- c (0.25, 0.75, 1.5, 3.5, 7.5)
    p_tro <- c (0.832, 0.119, 0.023, 0.005, 0.022)
    d_tro <- sum (d_tro * p_tro)

    d_work <- 0.474 * d_market + 0.19 * d_tro

    n_walk <- c (5, 15.5, 25.5, 35.5, 50.5, 80.5)
    p_walk <- c (0.64, 0.204, 0.062, 0.029, 0.034, 0.007)
    n_walk <- 7 * sum (n_walk * p_walk)

    d_walk_ref <- n_walk * d_tro + 2.5 * 0.474 * d_work + 2.5 * d_market

    d_work <- (0.474 * (1 + mode_incr)) * d_market +
        (0.19 * (1 + mode_incr)) * d_tro
    d_walk <- n_walk * (1 + mode_incr) * d_tro +
        2.5 * 0.474 * (1 + mode_incr) * d_work +
        2.5 * (1 + mode_incr) * d_market

    rr <- 0.114 * d_walk / d_walk_ref - 0.114
    accra_mortality <- accra_pop * mortality * rr
    data.frame (mode_shift = mode_incr,
                dist_ref = d_walk_ref,
                dist = d_walk,
                increase = d_walk / d_walk_ref - 1,
                rr = rr,
                d_mortality = accra_mortality)
}
mode_incr <- 0:10 / 100
accra_pop <- 2.27e6
x <- mode_shift (mode_incr, accra_pop, mortality)

par (mfrow = c (1, 2))
plot (x$mode_shift, x$increase, "l", col = "red", lwd = 2,
      xlab = "Relative mode shift",
      ylab = "Proportional increase in average trip length")
lines (c (0, 1), c (0, 1), col = "black", lty = 2)

plot (x$mode_shift, x$d_mortality, "l", col = "red", lwd = 2,
      xlab = "Relative mode shift",
      ylab = "Absolute reduction in all-cause mortality")
```

Increased walking also increases exposure to pollutants, as also quantified in
HEAT in terms of excess mortality arising through increased in active
transport. The following figure represents both the standard HEAT assumptions
of an increase in exposure (to PM2.5) of 1.6 for walking, and the specific
figure for Accra derived from the above literature of 1.31  (because background
pollution in Accra is so high to begin with).

```{r exposure, echo = FALSE}
exposure <- function (mode_incr = 0.01, accra_pop, mortality, rel_exp = "HEAT")
{
    x <- mode_shift (mode_incr, accra_pop, mortality)

    heat_vol_pm25_walk <- 1.37
    heat_vol_pm25_inactive <- 0.61

    # PM2.5 figures for Accra
    pm25bg <- 35
    pm25road <- pm25bg + 11
    rel_exp_walking <- pm25road / pm25bg # 1.31
    if (rel_exp == "HEAT")
        rel_exp_walking <- 1.6 # HEAT standard value

    # average weekly concentration for reference case:
    walk_time <- x$dist_ref / 5.3
    non_walk_time <- 24 * 7 - walk_time
    pm10_ref <- pm25bg * (walk_time * rel_exp_walking + non_walk_time) / (24 * 7)
    # modified average weekly concentration for scenario
    walk_time <- x$dist / 5.3
    non_walk_time <- 24 * 7 - walk_time
    pm10_scenario <- pm25bg * (walk_time * rel_exp_walking + non_walk_time) / (24 * 7)
    # capped at 50, but all well below here, so can be left as is

    d_exposure <- pm10_scenario - pm10_ref
    exposure <- accra_pop * mortality * d_exposure * 0.07 / 10

    data.frame (x,
                d_exposure = exposure,
                d_net = x$d_mortality - exposure)
}
x <- exposure (1:10 / 100, accra_pop, mortality)
x_accra <- exposure (1:10 / 100, accra_pop, mortality, rel_exp = "xx")
ylims <- range (c (x$d_mortality, x$d_exposure))
plot (x$mode_shift, x$d_mortality, "l", col = "grey", lwd = 2, ylim = ylims,
      xlab = "Relative mode shift to walking",
      ylab = "Absolute change in all-cause mortality")
lines (x$mode_shift, x$d_exposure, col = "grey", lwd = 2)
lines (x$mode_shift, x$d_net, lwd = 2)
lines (x$mode_shift, x_accra$d_net, lwd = 2, lty = 2)
legend ("topleft", lwd = 2, col = c ("grey", "grey", "black", "black"),
        bty = "n", lty = c (1, 2, 1, 2),
        legend = c ("Reduction due to physical activity",
                    "Increase due to pollutant exposure",
                    "Net reduction (HEAT default)",
                    "Net reduction (Accra-specific values)"))
```

## Spatial Scenarios

We adapted the procedure used to generate the above figures to provide
spatially explicit estimates of relative reduction in all-cause mortality, and
the net increase in all-cause mortality due to increased exposure. To do so, we
modelled vehicular flows through the city by presuming them to be proportional
to the betweenness centrality of the street network as weighted for vehicular
routing, using a time-based rather than distance-based routing algorithm.
Pedestrian flows were simulated as described in the accompanying manuscript
presenting our model of pedestrian flows throughout New York City. 

A given scenario of modal change was associated with a specific change in
average daily walking distance. We used the standard HEAT-assumed average
walking speed of 5.3 kilometres per hour to convert these values into total
daily durations of exposure to additional vehicular-borne pollutants. This
enabled an estimate of total daily exposure, through presuming the remainder of
the day to be spent exposed to the background rate of 35ug/m^3.

As described above, vehicular-borne emissions were presumed to disperse away
form the source according to a Gaussian dispersal kernel with a half-width of
200 metres. A flow of $f$ people along a particular street segment was then
modified under a given scenario with a modal increase in walking of $a$, to $f
(1 + a)$, and an associated increase in exposure proportional to the intensity
of diffused pollutants along that segment in relation to the background value.
We also incorporated our estimates of modal *decrease* in vehicular usage,
through assuming this to translate into a proportional decrease in overall
vehicular-borne emissions, and thus a concomitant decrease in expected lives
lost through increased exposure.

This procedure yielded the exposure layers presented in our Upthat tool, with
values along each street segment scaled such that the maximum value corresponds
to the expected *static* value according as presented in the above figure. It
is also noteworthy that the extremely high background levels of PM2.5
concentration in Accra mean that expected increases in mortality owing to
increases in exposure are actually very slight, generally less than one
additional death per year (further reduced by the presumed decrease in
vehicular emissions according to our models of modal shift). The spatially
explicit analyses are nevertheless uniquely powerful tools for highlight the
areas of the city in which exposure of pedestrians to vehicular-borne emissions
imposes the greatest health burden.


In contrast to the low values of expected additional mortality, the health
benefits of active transport translate to tens to hundreds of lives saved,
depending on overall modal shift. In a spatially explicit context, these
benefits are in direct proportion to expected pedestrian flows, and so are not
explicitly shown in our Upthat tool. The tool simply shows expected pedestrian
flows, with translation to expected lives saved depicted as part of our
scenario analyses.


