---
title: "Health Impacts of Scenarios"
output:
   #word_document: default
   #github_document
   #pdf_document:
   #  number_sections: yes
    html_document:
        toc: true
        toc_float: true
        number_sections: true
        theme: flatly
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  out.width = "100%",
  collapse = TRUE,
  comment = "#>",
  fig.path="figures/"
)
```
## Background - the WHO Health and Economic Assessment Tool (HEAT)


HEAT main website is [heatwalkingcycling.org](https://www.heatwalkingcycling.org).
The impact of physical activity is described
[here](https://www.heatwalkingcycling.org/#how_does_PA_work), the calculation
of mortality [here](https://www.heatwalkingcycling.org/#impact_calculation),
the effects of exposure to air pollution
[here](https://www.heatwalkingcycling.org/#how_does_AP_work), and default
values used in the tool are
[here](https://www.heatwalkingcycling.org/#generic_data).

### Physical Activity

The assumed relative reduction in risk for walking if 0.886, giving an effect
for walking a certain distance under a given scenario of

0.114 * d-scenario / d-reference

Risk reduction for walking is capped at 30%.

### Exposure

- Each increment of +10 ug/m^3 increases all-cause mortality risk by 1.07
- But only up to 50ug/m^3, beyond which no further increases occur
- Increase in exposure for walking is assumed to be a factor of 1.6

## Some Stats from, or useful for, Accra

Background mortality for HEAT is taken from the [WHO mortality
database](http://apps.who.int/healthinfo/statistics/mortality/whodpms). This is
difficult to access, so the table for all available counties is included here,
giving the following global average value:

```{r who-mortality}
x <- read.csv ("./who-mortality.csv")
names (x) <- c ("country", "population", "deaths", "remove")
x$remove <- NULL
x <- x [!is.na (x$deaths), ]
mortality <- mean (x$deaths / x$population)
message ("global mortality = ", mortality)
```

Population from wikipedia is 2.27 million

### Air pollution

An article on [air pollution in sub-Saharan African
cities](https://link.springer.com/article/10.1007%2Fs11869-013-0199-6) has an
appendix
[here](https://static-content.springer.com/esm/art%3A10.1007%2Fs11869-013-0199-6/MediaObjects/11869_2013_199_MOESM1_ESM.docx)
that includes figures for Accra from [this
paper](https://pubs.acs.org/doi/10.1021/es903276s). That estimates average
PM2.5 concentrations throughout the year *apart from* Dec-Jan, when dust blows
in from the Sahara, of 39-53ug/m^3 at roadside sites, and 30-70ug/m^3 at
residential sites. Across the whole year, average values at roadside sites were
80-108ug/m^3, and at residential sites 57-106ug/m^3.

The paper helpfully simply states that PM2.5 concentrations at roadside sites
were 8-14ug/m^3 higher at roadside than at residential sites. This suggests we
can assume a background value of 35ug/m^3, with road-borne emissions adding
another 14ug/m^3 at the highest vehicular intensities.

### Walking statistics

Taken from the data previously shared by WHO

Trips to school: 63.7% walk, and 15.3 take trotro
Trips to healthcare: 58% walk, 36% take taxis or trotros
Trips to work (male+female): 47.4% walk, 19% trotro

Accra distance to nearest market (km):

<1      | 1.1-2   | 2.1-3 | 3.1-6   | 6.1-10
--- | --- | --- | --- | ---
27.3    | 21.2    | 6.1   | 42.4    | 3.0

Trips per day on foot

0-10    | 11-20 | 21-30 | 31-40 | 41-60 | 61-100 | > 100
--- | --- | --- | --- | --- | --- | ---
64.0    | 20.4  | 6.2   | 2.9   | 3.4   | 0.7    | 0.0

Distance from home to trotro

0-0.5   | 0.6-1 | 1.1-2 | 2.1-5 | > 5
--- | --- | --- | --- | ---
83.2    | 11.9  | 2.3   | 0.5   | 2.2

Mode of transport to nearest food shop: walk = 93%
Mode of transport to nearest other shop: walk = 78.2%, trotro = 15.8%

These can be used to derive an average weekly distance walked, according to the
following steps. First, presume that trips to work are same as trips to markets:

```{r d_work}
d_market <- c (0.5, 1.5, 2.5, 4.5, 8.5)
p_market <- c (0.273, 0.212, 0.061, 0.424, 0.03)
d_market <- sum (d_market * p_market)

d_tro <- c (0.25, 0.75, 1.5, 3.5, 7.5)
p_tro <- c (0.832, 0.119, 0.023, 0.005, 0.022)
d_tro <- sum (d_tro * p_tro)

d_work <- 0.474 * d_market + 0.19 * d_tro
message ("Average distance to work = ", signif (d_work, 4), " km")
```

number of walking trips per day:
```{r n_walk}
n_walk <- c (5, 15.5, 25.5, 35.5, 50.5, 80.5)
p_walk <- c (0.64, 0.204, 0.062, 0.029, 0.034, 0.007)
n_walk <- sum (n_walk * p_walk)
message ("Number of walking trips per day = ", n_walk,
         " or per week = ", n_walk * 7)
n_walk <- n_walk * 7 # use weekly figures throughout
```

Then presume that weekly trips consist of 2-3 visits to a market, 2.5 trips to
work per person, and the remainder over distances typical of walks to trotro
stops. This gives a total weekly average distance of:
```{r d_walk_ref}
n_walk <- n_walk - 2.5 * 1.474 # (2.5 * 0.474 + 2.5 * 1 for market)
d_walk_ref <- n_walk * d_tro + 2.5 * 0.474 * d_work + 2.5 * d_market
message ("Average reference weekly distance walked = ",
         signif (d_walk_ref, 4), " km")
```
... and that is well beyond the HEAT-assumed baseline value of 14.84, but their
value is taken primarily from cities of the global North, not ones in which
almost all trips are made on foot.


## Scenarios

The following calculations give the effective increase in distance walked per
week in response to a relative increase of `mode_incr` - in other words, a mode
shift of that amount towards walking, and presumably away from other modes.
This is also translated, using standard HEAT values along with the global
average mortality rate, into an estimate of reduction in mortality in Accra.
```{r mode_shift}
x <- read.csv ("./who-mortality.csv")
names (x) <- c ("country", "population", "deaths", "remove")
x$remove <- NULL
x <- x [!is.na (x$deaths), ]
mortality <- mean (x$deaths / x$population)
mode_shift <- function (mode_incr = 0.01, accra_pop, mortality)
{
    d_market <- c (0.5, 1.5, 2.5, 4.5, 8.5)
    p_market <- c (0.273, 0.212, 0.061, 0.424, 0.03)
    d_market <- sum (d_market * p_market)

    d_tro <- c (0.25, 0.75, 1.5, 3.5, 7.5)
    p_tro <- c (0.832, 0.119, 0.023, 0.005, 0.022)
    d_tro <- sum (d_tro * p_tro)

    d_work <- 0.474 * d_market + 0.19 * d_tro

    n_walk <- c (5, 15.5, 25.5, 35.5, 50.5, 80.5)
    p_walk <- c (0.64, 0.204, 0.062, 0.029, 0.034, 0.007)
    n_walk <- 7 * sum (n_walk * p_walk)

    d_walk_ref <- n_walk * d_tro + 2.5 * 0.474 * d_work + 2.5 * d_market

    d_work <- (0.474 * (1 + mode_incr)) * d_market +
        (0.19 * (1 + mode_incr)) * d_tro
    d_walk <- n_walk * (1 + mode_incr) * d_tro +
        2.5 * 0.474 * (1 + mode_incr) * d_work +
        2.5 * (1 + mode_incr) * d_market

    rr <- 0.114 * d_walk / d_walk_ref - 0.114
    accra_mortality <- accra_pop * mortality * rr
    data.frame (mode_shift = mode_incr,
                dist_ref = d_walk_ref,
                dist = d_walk,
                increase = d_walk / d_walk_ref - 1,
                rr = rr,
                d_mortality = accra_mortality)
}
mode_incr <- 0:10 / 100
accra_pop <- 2.27e6
x <- mode_shift (mode_incr, accra_pop, mortality)

par (mfrow = c (1, 2))
plot (x$mode_shift, x$increase, "l", col = "red", lwd = 2,
      xlab = "Relative mode shift",
      ylab = "Proportional increase in average trip length")
lines (c (0, 1), c (0, 1), col = "black", lty = 2)

plot (x$mode_shift, x$d_mortality, "l", col = "red", lwd = 2,
      xlab = "Relative mode shift",
      ylab = "Absolute reduction in all-cause mortality")
```

Increased walking also increases exposure to pollutants, with the following
code quantifying those effects in the same way as HEAT. If `rel_exp` is
anything other than the default value, then the particular values for Accra are
used instead, which translate to a multiple for increased exposure during
walking of 1.31 instead of 1.6 (because background pollution in Accra is so
high to begin with).

```{r exposure}
exposure <- function (mode_incr = 0.01, accra_pop, mortality, rel_exp = "HEAT")
{
    x <- mode_shift (mode_incr, accra_pop, mortality)

    heat_vol_pm25_walk <- 1.37
    heat_vol_pm25_inactive <- 0.61

    # PM2.5 figures for Accra
    pm25bg <- 35
    pm25road <- pm25bg + 11
    rel_exp_walking <- pm25road / pm25bg # 1.31
    if (rel_exp == "HEAT")
        rel_exp_walking <- 1.6 # HEAT standard value

    # average weekly concentration for reference case:
    walk_time <- x$dist_ref / 5.3
    non_walk_time <- 24 * 7 - walk_time
    pm10_ref <- pm25bg * (walk_time * rel_exp_walking + non_walk_time) / (24 * 7)
    # modified average weekly concentration for scenario
    walk_time <- x$dist / 5.3
    non_walk_time <- 24 * 7 - walk_time
    pm10_scenario <- pm25bg * (walk_time * rel_exp_walking + non_walk_time) / (24 * 7)
    # capped at 50, but all well below here, so can be left as is

    d_exposure <- pm10_scenario - pm10_ref
    exposure <- accra_pop * mortality * d_exposure * 0.07 / 10

    data.frame (x,
                d_exposure = exposure,
                d_net = x$d_mortality - exposure)
}
x <- exposure (1:10 / 100, accra_pop, mortality)
x_accra <- exposure (1:10 / 100, accra_pop, mortality, rel_exp = "xx")
ylims <- range (c (x$d_mortality, x$d_exposure))
plot (x$mode_shift, x$d_mortality, "l", col = "grey", lwd = 2, ylim = ylims,
      xlab = "Relative mode shift to walking",
      ylab = "Absolute change in all-cause mortality")
lines (x$mode_shift, x$d_exposure, col = "grey", lwd = 2)
lines (x$mode_shift, x$d_net, lwd = 2)
lines (x$mode_shift, x_accra$d_net, lwd = 2, lty = 2)
legend ("topleft", lwd = 2, col = c ("grey", "grey", "black", "black"),
        bty = "n", lty = c (1, 2, 1, 2),
        legend = c ("Reduction due to physical activity",
                    "Increase due to pollutant exposure",
                    "Net reduction (HEAT default)",
                    "Net reduction (Accra-specific values)"))
```
