---
title: "Adaptation of UPTHAT to new locations"
output:
   #word_document: default
   #github_document
   #pdf_document:
   #  number_sections: yes
    html_document:
        toc: true
        toc_float: true
        number_sections: true
        theme: flatly
bibliography: ../ATF.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  out.width = "100%",
  collapse = TRUE,
  comment = "#>",
  fig.path="figures/"
)
```


This document describes the general principles of UPTHAT, with specific
instructions on how to adapt the analyses to new locations. The basis of UPTHAT
is detailed pedestrian flow model, able to be generated as described in the
following section. (This will soon be extended to include models of bicycle
flow, with their own accompanying adaptation section.) The second of the
following sections describes how to generate scenarios of mobility change for
new cities, and the third describes the steps necessary to translate such
scenarios into quantitative effect effects, largely building on the methodology
of the WHO's Health and Economic Assessment Tool (HEAT).

Almost all data inputs for UPTHAT can be derived from Open Street Map (OSM)
data. Not all locations are represented equally within OSM, and it may be that
a particular location is not sufficiently well represented for an UPTHAT model
to be accurately derived. The primary step that would need to be taken to
address any such data inadequacy would be to enhance the city's representation
in OSM, a step that would enhance any and all other present and future analyses
through enriching data on urban structure within what is arguably the world's
foremost publicly open and accessible database on urban form and structure.

Data within OSM are primarily structured in terms of so-called "key-value
pairs." A given entity is entered into the database as a series of one or more
coordinates (that is, longitude and latitude values), and described with an
arbitrary number of key-value pairs. Each key defines what is being described,
and each value provides the actual description. While anybody is free to enter
anything they desire into OSM, a set of well-documented guidelines define
standard or best practices for entering various urban structures. Streets,
paths, roads, and the like are, for example, all primarily defined with the key
of "highway", with associated values also generally defined by a relatively
small set of possible values.

# Modelling pedestrian flow

Our models of pedestrian flow extend from our initial calibration against
empirical pedestrian counts in New York City, which revealed the primary
generator of pedestrian trips to be public transport stops and stations
[@zacharias_pedestrian_2001]--most pedestrian trips involve people walking to
and from public transport. One of the primary initial requirements of UPTHAT
accordingly requires estimates of public transport usage, as described in the
following sub-section. Following that, we briefly describe the stages which
might be necessary to obtain an accurate representation of the street network
from OSM, before concluding this section with a description of building
structures and functions used in UPTHAT.

## Public Transportation

We presume for this section that a city has a bus system serving public
transportation needs (regardless of whether the buses themselves are public or
private), and that all (or at least most) bus lines and bus stops are
accurately represented in OSM. Train and tram lines can also be directly
incorporated, but are not yet described here in this initial incorporation of
our adaptation manual, because UPTHAT has not yet been adapted to any cities
which have these modes of public transport.

UPTHAT needs data both on entire bus routes, and on bus stops. OSM actually has
three primary kinds of data, representing single points, extended lines, and
complex higher-order relationships. Bus stops are almost always defined as
simple points, often associated with a key alone regardless of value. There are
two common ways within OSM of representing both bus stops and bus lines:

Representation | stop key | stop value | route key | route value
--- | --- | --- | --- | ---
1. | "bus" | "platform" | "type" | "route"
2. | "public_transport" | . | "route" | "bus"

Values defining stops for the second form are arbitrary, but are generally
specified as simple values of "yes".

Having established which form of representation a city has implemented, public
transport usage is estimated by extracting the OSM representation of the bus
network, and analysing the "centrality" of that network. This provides
a measure of how "central" each bus stop is within the entire urban network.
This metric is known to provide a reasonable proxy for public transport usage
[@derrible_network_2012; @jiang_station-based_2018]. 

A city may have its own estimates of, for example, bus stop usage, which may
provide more accurate estimates that our default approach of network analysis.
We have ensured that such data can be readily incorporated, but have not yet
automated this step.

## The street and way network

The next step is to extract the street and way network of a city. Street and
way networks within OSM are nearly complete for most of the world
[@barrington-leigh_worlds_2017], and this stage is generally very largely
automatic. There may nevertheless be local particularities that necessitate
manual processing. Open Street Map is often *very* up to date, and includes
numerous options for representing projects under construction, or in the
process of being demolished. Such temporary modifications to a city's street
network may have a noticeable effect of the output of UPTHAT, and require the
network to be inspected and potentially manually modified. For example, large
projects under construction may generally need to be incorporated within the
model in their anticipated state of completion. This generally just requires
identifying any such projects, and changing key-value pairs from, for example,
key = "highway", value = "construction", to key = "highway", value
= "motorway", or whatever appropriate value the project may represent on
completion.

One additional note on UPTHAT and the OSM network concerns pedestrian networks.
These may be less well represented in some cities that the street network
intended for motorized transport. Such cities may require additional
enhancement of the OSM representation of the pedestrianized aspects of their
way network, with such needs best address by modifying and contributing to OSM
itself, as described at the outset. We nevertheless note that all cities in
which the tool has been deployed have had very good representation of
pedestrian aspects of their
networks.

## Buildings

UPTHAT works by generating layers of pedestrian "flows" representing movement
for different categories of activities. These categories are in turn
represented in OSM by descriptions of buildings---for example, one category is
"health", for which movement is directed to or from all buildings with key
= "building" and value = "hospital" (along with several additional ways to
represent healthcare facilities in OSM). The more detailed the building
structure of a city within OSM is, the more detailed will be the resultant
UPTHAT model. While streets may be presumed to be universally well represented
within OSM, buildings may not, and many cities only have rudimentary
representations of their building structures. We have developed an alternative
approach for such cities, drawing on publicly accessible data from google, by
simply extracting all so-called "places of interest". Layers are then
constructed of movement to, from, and between all such points regardless of
actual purpose.

If the building structure of a city is sufficiently well represented, then the
stage of building extraction and processing can be fully automated. We are in
the process of developing a step to automatically quantify how well a city's
building structures are represented within OSM, so that the entire processing of
buildings within UPTHAT can be automated. Nevertheless, until this stage is
complete, cities need to minimally perform the single step of approximately
ascertaining how well their building structure is actually represented.

## Local Pedestrian Count Data

Once the above three types of data have been extracted and pre-processed, the
models of pedestrian movement between the various categories of origins and
destinations can be automatically calculated. These models should ideally be
calibrated to local circumstances, and in particularly to any data that may be
able to be provided by a city. The following represent the three main
possibilities of local data provision:

1. **No Data Able to be Provided** Cities which can provide no local data
   whatsoever can nevertheless receive a model of *relative* pedestrian flows
   throughout their cities. Such models will have an absolute scale provided
   through calibration with data from other locations, and so not necessarily
   appropriate to that particular city. In these cases, a city should presume
   both that numbers should be considered accurate only within some unknown
   multiplicative constant (such that real flows may be some $N$ times depicted
   flows, for some unknown value of $N$), and that spatial distributions will
   only approximate actual flows. Such models may nevertheless provide highly
   useful information, through providing a highly spatially detailed estimate
   of where people actually walk within a city.
2. **City Provides A Single Estimate of Actual Pedestrian Counts** More
   accurate results can be obtained where cities are to provide a single
   estimate of pedestrian counts. This can be either observed counts at
   a single location, or some city-wide estimate of, say, average numbers of
   daily pedestrian trips per person per day. This information is sufficient to
   ensure that the scales of pedestrian flows depicted in UPTHAT reflect actual
   pedestrian behaviour within a given city. The relative spatial distribution
   ought nevertheless be presumed to only approximately represent the actual
   distribution.
3. **City Provides Multiple Pedestrian Counts** Relative spatial distributions
   of pedestrian flows in UPTHAT can only be accurately estimated from multiple
   pedestrian counts within any city. This is because the models are formed by
   combining multiple layers, with the relative contribution of each layer only
   able to be estimated by comparing predictions within multiple observed
   pedestrian counts. Two pedestrian counts in a city will enable accurate
   combination of two flow layers; three counts will enable three layers to be
   combined, and so on. Our primary calibration procedure was developed for New
   York City, which has 114 count stations, and which enabled the accurate
   combination of up to 64 distinct layers. In short: the more empirical counts
   that a city has, the more accurate will be the resultant UPTHAT model.

# Pedestrian Exposure to Vehicular Emissions

The next stage in UPTHAT is to convert the model of pedestrian flows to a model
of exposure to vehicular emissions. This requires a model of vehicular flow
throughout a city. Unlike pedestrian flows, which are notoriously
unpredictable, vehicular flows are generally much more predictable, and are
generally well modelled by analysing the "centrality" of the vehicular network,
similar to the approach described above for public transport. We model
vehicular flows using time-based routing, which presumes drivers follow
fastest, yet not necessarily shortest, routes between origins and destinations.
Our models include all possible confounding factors such as differences in
speed limits on different kinds of roads, and time penalties for waiting at
traffic lights, and for turning across oncoming traffic. This stage can be
entirely automated, and cities will generally not need to manually intervene at
any stage in the generation of vehicular flow estimates, unless a city believes
that it already has a model more accurate than our own models, in which case
immediate substitution of one model for another is also very straightforward.

Having generated or otherwise obtained a model of vehicular flows, we then
presume vehicular emissions to disperse away from the source according to
a standard Gaussian dispersal process. UPTHAT requires specification of single
value describing the width of such dispersal, for which we currently use
a default value of 200 metres. The resultant layer of dispersed vehicular flows
then provides an estimate of the *relative* intensity of vehicular emissions
throughout a city. The exposure of pedestrians to these vehicular emissions can
then be derived by multiplying pedestrian flow intensities by these dispersed
vehicular emissions. We currently do this by multiplying absolute pedestrian
counts (in numbers per day) by vehicular emissions scaled to relative
intensities between zero and one. The result then provides an estimate for each
street segment of the relative exposure of pedestrians to the effects of
vehicular emissions.

The effects of such exposure on health are then evaluated by reference to our
scenario calculations, through using before-and-after calculations of a given
scenario as inputs to HEAT. For this prototype exercise, we have used values
provided by the WHO for Accra, Ghana.

## General parameters

We adopt the following general parameters directly from HEAT methodology. The
relative reduction in risk for walking is assumed to be 0.886, such that the effect
for walking a certain distance under a given scenario of
```}
0.114 * d-scenario / d-reference
```
where `d_reference` and `d_scenario` are respective distances walked under the
reference and scenario conditions. Risk reduction for walking is capped at 30%,
but does not approach that value in any of our models.

Each increment of exposure of +10 ug/m^3 (considered to refer to PM2.5
throughout, as in HEAT, with effects of other pollutants such as PM10 assumed
to be linearly related to effects of PM2.5) is assumed to increase all-cause
mortality risk by 1.07. Effects of increased exposure are presumed to cease above 
50ug/m^3. In contrast to HEAT, which assumes the increase in exposure for
walking to be a constant factor of 1.6, we use values derived from literature
specific to Accra, with spatial interpolation provided by our models as
described below.

In the absence of a specific general mortality rate for Accra, we simply
averaged values for all countries from the [WHO mortality
database](http://apps.who.int/healthinfo/statistics/mortality/whodpms). The
figures are included here, yielding a global average mortality rate as follows:
```{r who-mortality}
x <- read.csv ("./who-mortality.csv")
names (x) <- c ("country", "population", "deaths", "remove")
x$remove <- NULL
x <- x [!is.na (x$deaths), ]
mortality <- mean (x$deaths / x$population)
message ("global mortality = ", mortality)
```
The respective populations of Accra and Kathmandu are presumed to be is 2.27
and 1.74 million, as taken from from wikipedia.

## Accra-specific data 

### Air Pollution

An article on [air pollution in sub-Saharan African
cities](https://link.springer.com/article/10.1007%2Fs11869-013-0199-6) has an
appendix
[here](https://static-content.springer.com/esm/art%3A10.1007%2Fs11869-013-0199-6/MediaObjects/11869_2013_199_MOESM1_ESM.docx)
that includes figures for Accra from [this
paper](https://pubs.acs.org/doi/10.1021/es903276s). That estimates average
PM2.5 concentrations throughout the year *apart from* Dec-Jan, when dust blows
in from the Sahara, of 39-53ug/m^3 at roadside sites, and 30-70ug/m^3 at
residential sites. Across the whole year, average values at roadside sites were
80-108ug/m^3, and at residential sites 57-106ug/m^3.

The paper helpfully simply states that PM2.5 concentrations at roadside sites
were 8-14ug/m^3 higher at roadside than at residential sites. This suggests we
can assume a background value of 35ug/m^3, with road-borne emissions adding
another 14ug/m^3 at the highest vehicular intensities. The following analyses
assume these values, and allocate the additional 14ug/m^3 assumed to be caused
by vehicular pollution according to a standard model of automobile traffic
throughout the city. Vehicular emissions are presumed to disperse away from
direct linear (road-based) sources via a Gaussian dispersal model with an
assumed half-width of 200 metres. The following results are largely invariant
to precise choice of this dispersal width.

### Walking statistics

We derived estimates of average daily walking distance from  data previously
shared by WHO, as follows. Average distances walked to the nearest market (in
kilometres) were given as in the following table, which translates to an
average distance of 2.77 km:

<1      | 1.1-2   | 2.1-3 | 3.1-6   | 6.1-10
--- | --- | --- | --- | ---
27.3    | 21.2    | 6.1   | 42.4    | 3.0

Distances walked from home to the nearest bus (trotro) stop were similarly
tabulated as follows, translating to an average distance to trotro of 0.51 km:

0-0.5   | 0.6-1 | 1.1-2 | 2.1-5 | > 5
--- | --- | --- | --- | ---
83.2    | 11.9  | 2.3   | 0.5   | 2.2

We then estimated average distances walked to work from statistics for mode of
transport taken to work of 47.4% walking and 19% trotro. Walks to work were
assumed to reflect distances typical of the above walks to the nearest market,
giving a resultant mean distance to work of `d_work = (0.474 * d_work + 0.19
* d_trotro) / (0.474 + 0.19) = 2.12 km`

Numbers of trips per day was also tabulated as follows, translating to an
average of 11.2 walking trips per day.

0-10    | 11-20 | 21-30 | 31-40 | 41-60 | 61-100 | > 100
--- | --- | --- | --- | --- | --- | ---
64.0    | 20.4  | 6.2   | 2.9   | 3.4   | 0.7    | 0.0

This number of trips per day was converted to a weekly estimate of 78.8 trips.
Of these trips, an average of 2.5 per person per week were presumed to be trips
to work, and an equivalent number trips to the nearest markets (or other trips
over equivalent distances). The distances of one half of the remaining 73.8
trips were presumed to be typified by distances to trotro, and so have mean
values of 0.51 km, the remaining half to be double that distance. This yielded
an estimated average weekly distance walked of 66.3 km, or a daily value of
9.48 km. This figure may seem unusually high, especially compared with the
baseline value assumed in HEAT of 14.84 km *per week*, but it must also be
noted that 93% of all Accra residents walk to purchase food or other
necessities.

## Scenarios

The above figures can be translated into effective increase in distance walked
per week in response to a relative modal increases in walking. These increased
can then also be translated according to the standard HEAT values along with
the global average mortality rate, into an estimate of reduction in mortality
in Accra. The resultant figures are shown in the following graph.
```{r mode_shift, echo = FALSE}
x <- read.csv ("./who-mortality.csv")
names (x) <- c ("country", "population", "deaths", "remove")
x$remove <- NULL
x <- x [!is.na (x$deaths), ]
mortality <- mean (x$deaths / x$population)
mode_shift <- function (mode_incr = 0.01, accra_pop, mortality)
{
    d_market <- c (0.5, 1.5, 2.5, 4.5, 8.5)
    p_market <- c (0.273, 0.212, 0.061, 0.424, 0.03)
    d_market <- sum (d_market * p_market)

    d_tro <- c (0.25, 0.75, 1.5, 3.5, 7.5)
    p_tro <- c (0.832, 0.119, 0.023, 0.005, 0.022)
    d_tro <- sum (d_tro * p_tro)

    d_work <- 0.474 * d_market + 0.19 * d_tro

    n_walk <- c (5, 15.5, 25.5, 35.5, 50.5, 80.5)
    p_walk <- c (0.64, 0.204, 0.062, 0.029, 0.034, 0.007)
    n_walk <- 7 * sum (n_walk * p_walk)

    d_walk_ref <- n_walk * d_tro + 2.5 * 0.474 * d_work + 2.5 * d_market

    d_work <- (0.474 * (1 + mode_incr)) * d_market +
        (0.19 * (1 + mode_incr)) * d_tro
    d_walk <- n_walk * (1 + mode_incr) * d_tro +
        2.5 * 0.474 * (1 + mode_incr) * d_work +
        2.5 * (1 + mode_incr) * d_market

    rr <- 0.114 * d_walk / d_walk_ref - 0.114
    accra_mortality <- accra_pop * mortality * rr
    data.frame (mode_shift = mode_incr,
                dist_ref = d_walk_ref,
                dist = d_walk,
                increase = d_walk / d_walk_ref - 1,
                rr = rr,
                d_mortality = accra_mortality)
}
mode_incr <- 0:10 / 100
accra_pop <- 2.27e6
x <- mode_shift (mode_incr, accra_pop, mortality)

par (mfrow = c (1, 2))
plot (x$mode_shift, x$increase, "l", col = "red", lwd = 2,
      xlab = "Relative mode shift",
      ylab = "Proportional increase in average trip length")
lines (c (0, 1), c (0, 1), col = "black", lty = 2)

plot (x$mode_shift, x$d_mortality, "l", col = "red", lwd = 2,
      xlab = "Relative mode shift",
      ylab = "Absolute reduction in all-cause mortality")
```

Increased walking also increases exposure to pollutants, as also quantified in
HEAT in terms of excess mortality arising through increased in active
transport. The following figure represents both the standard HEAT assumptions
of an increase in exposure (to PM2.5) of 1.6 for walking, and the specific
figure for Accra derived from the above literature of 1.31  (because background
pollution in Accra is so high to begin with).

```{r exposure, echo = FALSE}
exposure <- function (mode_incr = 0.01, accra_pop, mortality, rel_exp = "HEAT")
{
    x <- mode_shift (mode_incr, accra_pop, mortality)

    heat_vol_pm25_walk <- 1.37
    heat_vol_pm25_inactive <- 0.61

    # PM2.5 figures for Accra
    pm25bg <- 35
    pm25road <- pm25bg + 11
    rel_exp_walking <- pm25road / pm25bg # 1.31
    if (rel_exp == "HEAT")
        rel_exp_walking <- 1.6 # HEAT standard value

    # average weekly concentration for reference case:
    walk_time <- x$dist_ref / 5.3
    non_walk_time <- 24 * 7 - walk_time
    pm10_ref <- pm25bg * (walk_time * rel_exp_walking + non_walk_time) / (24 * 7)
    # modified average weekly concentration for scenario
    walk_time <- x$dist / 5.3
    non_walk_time <- 24 * 7 - walk_time
    pm10_scenario <- pm25bg * (walk_time * rel_exp_walking + non_walk_time) / (24 * 7)
    # capped at 50, but all well below here, so can be left as is

    d_exposure <- pm10_scenario - pm10_ref
    exposure <- accra_pop * mortality * d_exposure * 0.07 / 10

    data.frame (x,
                d_exposure = exposure,
                d_net = x$d_mortality - exposure)
}
x <- exposure (1:10 / 100, accra_pop, mortality)
x_accra <- exposure (1:10 / 100, accra_pop, mortality, rel_exp = "xx")
ylims <- range (c (x$d_mortality, x$d_exposure))
plot (x$mode_shift, x$d_mortality, "l", col = "grey", lwd = 2, ylim = ylims,
      xlab = "Relative mode shift to walking",
      ylab = "Absolute change in all-cause mortality")
lines (x$mode_shift, x$d_exposure, col = "grey", lwd = 2)
lines (x$mode_shift, x$d_net, lwd = 2)
lines (x$mode_shift, x_accra$d_net, lwd = 2, lty = 2)
legend ("topleft", lwd = 2, col = c ("grey", "grey", "black", "black"),
        bty = "n", lty = c (1, 2, 1, 2),
        legend = c ("Reduction due to physical activity",
                    "Increase due to pollutant exposure",
                    "Net reduction (HEAT default)",
                    "Net reduction (Accra-specific values)"))
```

These idealised scenarios can then be given specific context through applying
our procedures described in our separate "Scenarios" document to a particular
city, in order to generate locally realistic scenarios of response to such
measures as the addition of bus stops, or the restriction of motorized vehicles
in the inner city. The outputs of our scenario calculations will then provide
absolute values for expected increases in numbers and/or lengths of walking
trips, along with associated decreases in numbers and/or lengths of trips made
in private motorized vehicles.

## Spatial Scenarios

We adapted the procedure used to generate the above figures to provide
spatially explicit estimates of relative reduction in all-cause mortality, and
the net increase in all-cause mortality due to increased exposure. To do so, we
modelled vehicular flows through the city by presuming them to be proportional
to the betweenness centrality of the street network as weighted for vehicular
routing, using a time-based rather than distance-based routing algorithm.
Pedestrian flows were simulated as described in the accompanying manuscript
presenting our model of pedestrian flows throughout New York City. 

A given scenario of modal change was associated with a specific change in
average daily walking distance. We used the standard HEAT-assumed average
walking speed of 5.3 kilometres per hour to convert these values into total
daily durations of exposure to additional vehicular-borne pollutants. This
enabled an estimate of total daily exposure, through presuming the remainder of
the day to be spent exposed to the background rate of 35ug/m^3.

As described above, vehicular-borne emissions were presumed to disperse away
form the source according to a Gaussian dispersal kernel with a half-width of
200 metres. A flow of $f$ people along a particular street segment was then
modified under a given scenario with a modal increase in walking of $a$, to $f
(1 + a)$, and an associated increase in exposure proportional to the intensity
of diffused pollutants along that segment in relation to the background value.
We also incorporated our estimates of modal *decrease* in vehicular usage,
through assuming this to translate into a proportional decrease in overall
vehicular-borne emissions, and thus a concomitant decrease in expected lives
lost through increased exposure.

This procedure yielded the exposure layers presented in our Upthat tool, with
values along each street segment scaled such that the maximum value corresponds
to the expected *static* value according as presented in the above figure. It
is also noteworthy that the extremely high background levels of PM2.5
concentration in Accra mean that expected increases in mortality owing to
increases in exposure are actually very slight, generally less than one
additional death per year (further reduced by the presumed decrease in
vehicular emissions according to our models of modal shift). The spatially
explicit analyses are nevertheless uniquely powerful tools for highlight the
areas of the city in which exposure of pedestrians to vehicular-borne emissions
imposes the greatest health burden.


In contrast to the low values of expected additional mortality, the health
benefits of active transport translate to tens to hundreds of lives saved,
depending on overall modal shift. In a spatially explicit context, these
benefits are in direct proportion to expected pedestrian flows, and so are not
explicitly shown in our Upthat tool. The tool simply shows expected pedestrian
flows, with translation to expected lives saved depicted as part of our
scenario analyses.


# References
